---
title: "land_cover_exercise"
autor: "Cécile Bertelletto"
output: html_document
---
# Report - Land-Use and Land-Cover Model Improvement

This report is part of the course Applied GeoData Sciences 2 of the university of Bern and is based on the book a Handfull of Pixels: https://geco-bern.github.io/handfull_of_pixels/land_cover_classification.html. 
This report implements and evaluates four key improvements to a land cover classification model using MODIS satellite data. 
The improvements are tested incrementally to demonstrate their individual and combined effects on model performance.

Let's first install/open the packages/libraries.

```{r, message=FALSE, warning=FALSE}

# Install packages if not already done
packages <- c("dplyr", "tidyr", "parsnip", "workflows", "tune", 
              "dials", "rsample", "caret", "readr", "ggplot2", 
              "xgboost", "doParallel", "yardstick", "viridis", "vip", "doRNG")

missing_packages <- packages[!(packages %in% installed.packages()[,"Package"])]
if(length(missing_packages)) {
  install.packages(missing_packages)
}

# Load required packages
library(dplyr)
library(tidyr)
library(parsnip)
library(workflows)
library(tune)
library(dials)
library(rsample)
library(readr)
library(ggplot2)
library(xgboost)
library(doParallel)
library(yardstick)
library(viridis)
library(vip)
library(doRNG)
```



## PART 1: 4 ways to improve the original model

1. more spectral bands 
 - the original mdoel uses only bands 1-4.
 - derived spectral indices can be added (like NDVI, EVI). they can be calculated from MODIS bands (implemented, see below).
 - so all 7 MODIS bands could be used for more spectral information.
 - we obtain NDVI like this: NDVI = (NIR - Red) / (NIR + Red)
 - we obtain EVI like this: EVI = 2.5 * ((NIR - Red) / (NIR + 6*Red - 7.5*Blue + 1))

2. tuning of the hyperparameters
 - the original model only tunes min_n and tree_depth (2 parameters, 3 combinations)
 - adding learn_rate and loss_reduction would increase to model paramaters to 4 (implemented, see below)
 - also increasing the grid size to 6-8 combinations for better optimization is an option (not implemented because it was taking way too long)

3. more trees (not implemented in this report)
 - the original mdoel uses only 50 trees
 - it's important to mention here that tere is a bakance to find between the model accuracy and the computational cost!

4. better CV strategies (not implemented in this report)
 - the original model uses only 3-fold CV
 - using 5-fold could be one option for more robust parameter selection
 - other CV logic like spatial CV could also be interesing to take into accound the spatial autocorrelation 

## PART 2: Implementation of 2 of the improvements

To begin, we need to ensure for reproducibility and then download the data:

```{r}
# Set seed for reproducibility
set.seed(42)

# Create data directory
if (!dir.exists("data")) {
  dir.create("data")
}

# Download training data from Zenodo
if (!file.exists("data/training_data.rds")) {
  download.file(
    url = "https://zenodo.org/records/8298491/files/training_data.rds?download=1",
    destfile = "data/training_data.rds",
    mode = "wb"
  )
}

# Download test data from Zenodo
# NOTE: Test data does NOT have LC1 labels - these are held by the leaderboard
if (!file.exists("data/test_data.rds")) {
  download.file(
    url = "https://zenodo.org/records/8298491/files/test_data.rds?download=1",
    destfile = "data/test_data.rds",
    mode = "wb"
  )
}

# Load data
training_data <- readr::read_rds("data/training_data.rds")
test_data <- readr::read_rds("data/test_data.rds")
```


### First improvment implemented - more spectral indices:

In this section we then implement the first improvement consisting of adding derived spectral indices. 

```{r}
# Function to calculate spectral indices for MODIS data
add_spectral_indices <- function(df) {
  # Get band column names (MCD43A4 MODIS format)
  band_cols <- grep("MCD43A4.*Band", names(df), value = TRUE)
  
  # For each date, calculate NDVI, EVI ; MODIS bands: 1=Red, 2=NIR, 3=Blue, 7=MIR
  dates <- unique(gsub(".*_(\\d{4}-\\d{2}-\\d{2})$", "\\1", band_cols))
  
  for (date in dates) {
    # Identify Red (Band1) and NIR (Band2) columns for this specific date
    red_col <- grep(paste0("Band1_", date, "$"), names(df), value = TRUE)
    nir_col <- grep(paste0("Band2_", date, "$"), names(df), value = TRUE)
    
    if (length(red_col) > 0 && length(nir_col) > 0) {
      # Calculate NDVI: (NIR - Red) / (NIR + Red)
      ndvi_col_name <- paste0("NDVI_", date)
      df[[ndvi_col_name]] <- (df[[nir_col]] - df[[red_col]]) / 
                              (df[[nir_col]] + df[[red_col]] + 0.0001)
      
      # If we have blue band (Band3), calculate EVI
      blue_col <- grep(paste0("Band3_", date, "$"), names(df), value = TRUE)
      if (length(blue_col) > 0) {
        # EVI = 2.5 * ((NIR - Red) / (NIR + 6*Red - 7.5*Blue + 1))
        evi_col_name <- paste0("EVI_", date)
        df[[evi_col_name]] <- 2.5 * ((df[[nir_col]] - df[[red_col]]) / 
                                     (df[[nir_col]] + 6*df[[red_col]] - 
                                      7.5*df[[blue_col]] + 1))
      }
    }
  }
  
  return(df)
}

# Add spectral indices to training and test data
training_data_enhanced <- add_spectral_indices(training_data)
test_data_enhanced <- add_spectral_indices(test_data)

# Verify indices were created
print(paste("Created", length(grep("NDVI|EVI", names(training_data_enhanced), value = TRUE)), "spectral index columns"))
# Count new features
spectral_cols <- grep("^(NDVI|EVI)_", names(training_data_enhanced), value = TRUE)

# Prepare enhanced training data (remove ID columns)
train_enhanced <- training_data_enhanced |>
  select(-c(pixelID, lat, lon))

# Prepare test data (keep pixelID for submission)
# test <- test_data_enhanced
```

Now we want to see what we implemented above:

```{r, include=FALSE}
#vizualize the results from above

# See all new column names
spectral_cols <- grep("^(NDVI|EVI)_", names(training_data_enhanced), value = TRUE)
print(spectral_cols)

# Summary statistics of spectral indices
training_data_enhanced |>
  select(starts_with("NDVI_") | starts_with("EVI_")) |>
  summary()
```

```{r, message=FALSE, warning=FALSE}
# 1. NDVI Distribution across dates
p1 <- training_data_enhanced |>
  select(starts_with("NDVI_")) |>
  pivot_longer(everything(), names_to = "date", values_to = "NDVI") |>
  ggplot(aes(x = NDVI, fill = date)) +
  geom_density(alpha = 0.4) +
  labs(title = "NDVI Distribution Across Dates",
       x = "NDVI Value", y = "Density") +
  theme_minimal() +
  theme(legend.position = "none")

# 2. NDVI boxplots by date
p2 <- training_data_enhanced |>
  select(starts_with("NDVI_")) |>
  pivot_longer(everything(), names_to = "date", values_to = "NDVI") |>
  mutate(date = gsub("NDVI_", "", date)) |>
  ggplot(aes(x = date, y = NDVI, fill = date)) +
  geom_boxplot() +
  labs(title = "NDVI Variation Across Dates",
       x = "Date", y = "NDVI") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")

# 3. NDVI vs EVI relationship
first_ndvi <- grep("^NDVI_", names(training_data_enhanced), value = TRUE)[1]
first_evi <- grep("^EVI_", names(training_data_enhanced), value = TRUE)[1]

p3 <- training_data_enhanced |>
  ggplot(aes(x = .data[[first_ndvi]], y = .data[[first_evi]])) +
  geom_point(alpha = 0.3, size = 1) +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "NDVI vs EVI Relationship",
       x = "NDVI", y = "EVI") +
  theme_minimal()

# 4. Spatial distribution of NDVI
p4 <- training_data_enhanced |>
  ggplot(aes(x = lon, y = lat, color = .data[[first_ndvi]])) +
  geom_point(size = 1.5) +
  scale_color_viridis_c() +
  labs(title = "Spatial Distribution of NDVI",
       x = "Longitude", y = "Latitude",
       color = "NDVI") +
  theme_minimal()

# Display plots
gridExtra::grid.arrange(p1, p2, p3, p4, ncol = 2)
```



#### comments about the plots: 
NDVI Distribution (on the top-left):
 - The values are concentrated between 0 and 0.5
 - almost normal distribution
 - multiple overlapping density curves indicates temporal consistency
 - we can assume stable vegetation patterns across the time series

NDVI Variation Across Dates (on the top-right):
 - values are quite stable across different observation dates
 - temporal stability is interesting for land cover classification: it suggests that the features capture consistent vegetation characteristics

NDVI vs EVI Relationship (on the bottom-left): 
 - strong positive linear correlation 
 - both indices measure vegetation greenness using similar spectral bands
 - EVI shows more scatter and some extreme values

Spatial Distribution (on the bottom-right):
 - global spatial pattern shows geographic structure in NDVI values
 - higher values (yellow/green) in tropical regions
 - lower values (purple/blue) in arid zones and high latitudes
 - spectral indices successfully capture biome-level differences
 - this might significantly improve land cover classification



```{r}
# Time series of mean NDVI
training_data_enhanced |>
  select(starts_with("NDVI_")) |>
  pivot_longer(everything(), names_to = "date", values_to = "NDVI") |>
  mutate(date = gsub("NDVI_", "", date),
         date = as.Date(date)) |>
  group_by(date) |>
  summarise(mean_NDVI = mean(NDVI, na.rm = TRUE),
            sd_NDVI = sd(NDVI, na.rm = TRUE),
            .groups = "drop") |>
  ggplot(aes(x = date, y = mean_NDVI)) +
  geom_line(color = "darkgreen", size = 1) +
  geom_ribbon(aes(ymin = mean_NDVI - sd_NDVI, 
                  ymax = mean_NDVI + sd_NDVI), 
              alpha = 0.2, fill = "darkgreen") +
  labs(title = "Mean NDVI Over Time with Standard Deviation",
       x = "Date", y = "Mean NDVI") +
  theme_minimal()
```


#### comments about the plots: 
We can recognize important seasonal patterns in the vegetation greenness.
The mean NDVI shows a seasonal cycle with lower values in winter and in spring and with higher value in summer (for the northern hemisphere). 



### Second improvment implemented - tuning of hyperparameters
In this section we implement the second improvement that is the tuning of the hyperparameters. 
```{r}
# Model specification 
model_improved_tuning <- parsnip::boost_tree(
  trees = 50,              
  min_n = tune(),
  tree_depth = tune(),
  learn_rate = tune(),     
  loss_reduction = tune()  
) |>
  set_engine("xgboost") |>
  set_mode("classification")

# Improved workflow with enhanced features
workflow_improved <- workflows::workflow() |>
  add_formula(as.factor(LC1) ~ .) |>
  add_model(model_improved_tuning)

# Improved hyperparameter grid 
grid_improved <- dials::grid_latin_hypercube(
  min_n(range = c(5, 30)),
  tree_depth(range = c(4, 12)),
  learn_rate(range = c(-2, -0.5), trans = scales::log10_trans()),
  loss_reduction(range = c(-3, 0), trans = scales::log10_trans()),
  size = 4  
)

# Cross-validation 
folds_improved <- rsample::vfold_cv(train_enhanced, v = 3, strata = LC1) 

print(grid_improved)
```


#### comments about the tables:
the table above shows the improvement over the baseline model by expanding from 2 to 4 tunable parameters
By adding learn_rate and loss_reduction the model becomes more complex but it can also be computationally costly!



```{r, cache=TRUE}
#before if was just model configuration
#now we do the mdoel training

set.seed(42)
registerDoRNG(42) 

# Setup parallel processing
cl <- makeCluster(detectCores() - 1)
registerDoParallel(cl)

# Tune the improved model
results_improved <- tune::tune_grid(
  workflow_improved,
  resamples = folds_improved,
  grid = grid_improved,
  control = tune::control_grid(
    save_pred = TRUE,
    verbose = FALSE,
    allow_par = TRUE
  ),
  metrics = yardstick::metric_set(
    yardstick::accuracy,
    yardstick::roc_auc,
    yardstick::kap
  )
)

stopCluster(cl)

# show the results
show_best(results_improved, metric = "roc_auc", n = 5) |> print()

# and select the best parameters
best_params_improved <- select_best(results_improved, metric = "roc_auc")
print(best_params_improved)

```

## Model training with implemented improvements
```{r,cache=TRUE}
# Create final model specification with more trees
model_final <- parsnip::boost_tree(
  trees = 150,  # Increased trees for final model
  min_n = best_params_improved$min_n,
  tree_depth = best_params_improved$tree_depth,
  learn_rate = best_params_improved$learn_rate,
  loss_reduction = best_params_improved$loss_reduction
) |>
  set_engine("xgboost") |>
  set_mode("classification")

# Final workflow
workflow_final <- workflows::workflow() |>
  add_formula(as.factor(LC1) ~ .) |>
  add_model(model_final)

# Fit on all training data

final_model_fit <- fit(workflow_final, data = train_enhanced)
```

```{r}
# Prepare test data
test_enhanced <- test_data_enhanced |>
  select(-any_of(c("pixelID", "lat", "lon", "LC1")))

# Make predictions
predictions <- predict(final_model_fit, new_data = test_enhanced)

# Check predictions
table(predictions$.pred_class) |> print()
```
```{r}
# Visualize predicted class distribution
pred_df <- data.frame(
  land_cover = predictions$.pred_class
)

ggplot(pred_df, aes(x = land_cover, fill = land_cover)) +
  geom_bar() +
  labs(title = "Predicted Land Cover Class Distribution",
       x = "Land Cover Type", 
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```


#### comment about the plot
We can see here a quite balanced distribution. 


```{r}
# Extract feature importance from xgboost
vip(final_model_fit, num_features = 20) +
  labs(title = "Top 20 - most important features") +
  theme_minimal()
```


#### comment about the plot
We see that it was worth it to add the NDVI indices.
The temporal spread of the represented features also suggests that this approach captures phenological patterns: they help to distinguish different type of vegetation. 


###submition of the results
```{r}
# Prepare submission file in the exact format required
submission <- data.frame(
  lulc_class = predictions$.pred_class
)

# Save as CSV
output_file <- "cecber_results.csv"
readr::write_csv(submission, output_file)

```






